---
Title: "Mixed Effects Analysis of Longitudinal Cholesterol Data"
Author: "Sushma Vuppala"
Date: "2025-01-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Background

Cyclosporine A (CsA) is an immunosuppressive agent widely used in renal transplant recipients, but it is associated with hyperlipidemia — significantly raising total cholesterol, LDL cholesterol, and apoprotein B levels. In 1994, Tacrolimus (FK506) received FDA approval as an alternative immunosuppressant shown to have less adverse effect on post-transplant lipid profiles.

This randomized controlled trial (RCT) investigated whether it was safe to convert stable renal transplant recipients (at least 12 months post-transplant, total cholesterol > 240 mg/dL) from CsA to FK506, and whether conversion would result in a clinically significant reduction in total cholesterol (defined as a drop of at least 40 mg/dL). Cholesterol was measured at Baseline, Month 1, Month 3, and Month 6.

---

## Step 1: Load Required Libraries

```{r libraries}
library(dplyr)       # Data manipulation
library(reshape2)    # Wide to long transformation
library(lme4)        # Linear mixed model (lmer) — used to confirm model justification
library(lmerTest)    # F-tests with Satterthwaite degrees of freedom for lmer
library(nlme)        # Linear mixed model (lme) — used for covariance structure selection
library(emmeans)     # Estimated marginal means and post-hoc comparisons
library(performance) # ICC calculation
library(ggplot2)     # Data visualization
```

---

## Step 2: Import Data

```{r import}
setwd("/Users/sushmavuppala/Documents/RCT Project")

data <- read.csv("RCT_Data.csv")

head(data)
```

---

## Step 3: Data Inspection and Cleaning

```{r cleaning}
# Review structure and distributions
str(data)
summary(data)

# Check for missing values and duplicate records
sapply(data, function(x) sum(is.na(x)))
sum(duplicated(data))

# Convert participant identifier and treatment group to categorical factors
data$id    <- factor(data$id)
data$group <- factor(data$group,
                     levels = c("1", "2"),
                     labels = c("FK506 (Treatment)", "CsA (Control)"))

str(data)
```

The dataset contained 54 participants and 6 variables: participant identifier (`id`), treatment group (`group`: FK506 vs. CsA), and total cholesterol measurements at Baseline, Month 1, Month 3, and Month 6. No missing values or duplicate records were identified. The dataset was balanced, with 27 participants per group.

---

## Step 4: Reshape to Long Format

```{r reshape}
# Convert from wide format (one row per subject) to long format (one row per observation)
# This is required for longitudinal mixed-effects modeling
long_data <- melt(
  data,
  id.vars       = c("id", "group"),
  measure.vars  = c("chol0", "chol1", "chol3", "chol6"),
  variable.name = "time",
  value.name    = "cholesterol"
)

head(long_data)
```

---

## Step 5: Data Transformation 

```{r transformation}
# Relabel time points with clinically meaningful labels
long_data$time <- factor(
  long_data$time,
  levels = c("chol0", "chol1", "chol3", "chol6"),
  labels = c("Baseline", "Month 1", "Month 3", "Month 6")
)

# Create numeric time variable using true month values (0, 1, 3, 6)
# This correctly reflects the unequal spacing between visits
# Used by SP(EXP) and AR(1) covariance structures in Step 7
long_data$time_numeric <- ifelse(long_data$time == "Baseline", 0,
                          ifelse(long_data$time == "Month 1",  1,
                          ifelse(long_data$time == "Month 3",  3, 6)))

# Create consecutive integer time variable (1, 2, 3, 4)
# Required specifically by corSymm (Unstructured) which needs consecutive integers
long_data$time_int <- as.integer(long_data$time)
```

---

## Step 6: Data Validation

```{r validation}
# Verify final structure
str(long_data)

# Confirm no missing values after transformation
sum(is.na(long_data$cholesterol))

# Confirm balanced design — equal observations per group and time point
table(long_data$group, long_data$time)
```

After reshaping, the dataset contained 216 observations (54 participants × 4 time points). The design was fully balanced with 27 observations per group per time point and no missing data.

---

## Step 7: Exploratory Data Analysis

```{r eda-summary}
# Summary statistics with 95% confidence intervals for each group × time combination
eda_summary <- long_data %>%
  group_by(group, time) %>%
  summarise(
    n        = n(),
    mean     = mean(cholesterol, na.rm = TRUE),
    sd       = sd(cholesterol, na.rm = TRUE),
    se       = sd / sqrt(n),
    ci_lower = mean - 1.96 * se,
    ci_upper = mean + 1.96 * se,
    .groups  = "drop"
  )

eda_summary
```


```{r eda-plot}
# Visualize mean cholesterol trajectories over time by treatment group
ggplot(long_data, aes(x = time, y = cholesterol, color = group, group = group)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  scale_color_manual(values = c("FK506 (Treatment)" = "#1B6CA8",
                                "CsA (Control)"     = "#C43C3C")) +
  labs(
    title = "Mean Total Cholesterol Over Time by Treatment Group",
    x     = "Time Point",
    y     = "Total Cholesterol (mg/dL)",
    color = "Treatment Group"
  ) +
  theme_minimal() +
  theme(
    plot.title      = element_text(hjust = 0.5),
    legend.position = "bottom"
)
```

The FK506 group showed a pronounced reduction in total cholesterol from Baseline (~290 mg/dL) to Month 1 (~218 mg/dL), exceeding the pre-specified clinically significant threshold of 40 mg/dL. The CsA group showed a smaller, more gradual decline over the same period. This divergence in trajectories suggests a differential treatment effect, which is formally evaluated in the mixed-effects model below.

---

## Step 8: Linear Mixed Model — Confirming Model Justification (lmer)

Before selecting a covariance structure, we first fit a linear mixed model using `lmer()` to confirm that a mixed-effects approach is statistically justified. The model includes treatment group, time, and their interaction as fixed effects, and a subject-specific random intercept to account for the correlation introduced by repeated measurements on the same participant.

At this stage the model assumes compound symmetry — that the correlation between any two time points is equal regardless of how far apart they are. This assumption will be formally evaluated and relaxed in Step 9.

```{r lmer-model}
model_lmer <- lmer(cholesterol ~ group * time + (1 | id), data = long_data)

summary(model_lmer)
```

```{r lmer-anova}
# Type III F-tests for fixed effects using Satterthwaite degrees of freedom
anova(model_lmer)
```

```{r lmer-icc}
# Intraclass Correlation Coefficient (ICC)
# Quantifies the proportion of total variance attributable to between-subject differences
# ICC > 0.1 generally justifies use of a mixed model
icc(model_lmer)

# Likelihood ratio test for the random intercept
# Compares model with vs. without random intercept to confirm it is needed
rand(model_lmer)
```

The group × time interaction was highly significant (F = 8.84, p < 0.001), confirming that cholesterol changed differently over time between the FK506 and CsA groups. The ICC of 0.506 indicates that 50% of total variance is attributable to stable between-subject differences, strongly justifying the mixed-effects approach. The random intercept was highly significant (LRT p = 3.3e-15), confirming that 
participants differ meaningfully in their baseline cholesterol levels.

---

## Step 9: Covariance Structure Selection (lme)

The `lmer` model above assumed compound symmetry — equal correlation between all pairs of time points. In longitudinal studies, this assumption is often violated because measurements taken closer together in time tend to be more strongly correlated than measurements further apart. We therefore use `lme()` from the `nlme` package, which provides the `correlation =` argument to explicitly specify and compare 
alternative within-subject covariance structures.

Note: `lme()` fits the same linear mixed model as `lmer()`. The key difference is that `lme()` allows the residual correlation structure to be modeled flexibly, which `lmer()` does not support.

Five candidate covariance structures are compared using AIC and BIC:

```{r covariance-models}
# VC: Variance Components
# Assumes complete independence between time points — no within-subject correlation
model_VC <- lme(cholesterol ~ group * time,
                random = ~1 | id,
                data = long_data)

# CS: Compound Symmetry
# Assumes all pairs of time points share the same correlation
# Equivalent to the lmer assumption in Step 7
model_CS <- lme(cholesterol ~ group * time,
                random = ~1 | id,
                correlation = corCompSymm(form = ~time_numeric | id),
                data = long_data)

# AR(1): First-Order Autoregressive
# Correlation decays exponentially with time — closer visits more correlated
# Assumes equal spacing between visits, which does not hold here (0,1,3,6 months)
model_AR <- lme(cholesterol ~ group * time,
                random = ~1 | id,
                correlation = corAR1(form = ~time_numeric | id),
                data = long_data)

# SP(EXP): Spatial Exponential
# Correlation decays with time and correctly accounts for unequal visit spacing
# Most appropriate structure given visits at months 0, 1, 3, and 6
model_SP <- lme(cholesterol ~ group * time,
                random = ~1 | id,
                correlation = corSpatial(form = ~time_numeric | id, type = "exp"),
                data = long_data)

# UN: Unstructured
# Each pair of time points has its own freely estimated correlation
# Most flexible but most parameter-intensive
# Note: corSymm requires consecutive integers (1,2,3,4) — uses time_int not time_numeric
model_UN <- lme(cholesterol ~ group * time,
                random = ~1 | id,
                correlation = corSymm(form = ~time_int | id),
                weights = varIdent(form = ~1 | time),
                data = long_data)
```



```{r model-comparison}
# Compare all five structures using AIC and BIC
# Lower values indicate better fit, penalized for model complexity
model_comparison <- data.frame(
  Model       = c("VC", "CS", "AR(1)", "SP(EXP)", "UN"),
  Description = c("Independence", "Equal correlations",
                  "Decaying correlation (equal spacing)",
                  "Decaying correlation (unequal spacing)",
                  "Freely estimated correlations"),
  AIC = c(AIC(model_VC), AIC(model_CS), AIC(model_AR),
          AIC(model_SP), AIC(model_UN)),
  BIC = c(BIC(model_VC), BIC(model_CS), BIC(model_AR),
          BIC(model_SP), BIC(model_UN))
)

model_comparison
```

The spatial exponential model SP(EXP) accounts for the unequal spacing between visits (0, 1, 3, 6 months) and is therefore the most theoretically appropriate structure for this study design. AR(1) assumes equal spacing and is misspecified here. The model with the lowest AIC was selected as the final model.



```{r final-model}
# Final model selected based on AIC/BIC and clinical appropriateness
final_model <- model_SP

summary(final_model)
anova(final_model)
```

The final spatial exponential model confirmed a significant group × time interaction (F = 11.55, p < 0.001), indicating that total cholesterol changed differently over time between the FK506 and CsA groups. The main effects of group (F = 29.07, p < 0.001) and time (F = 43.96, p < 0.001) were also highly significant.

---

## Step 10: Estimated Marginal Means and Post-Hoc Comparisons

Estimated marginal means (LS Means) are derived from the final model for each group × time combination, adjusted for all other terms. Pairwise comparisons are adjusted using the Tukey method to control the family-wise error rate across multiple comparisons.

```{r emmeans}
# Extract LS Means from final model
emm <- emmeans(final_model, ~ group * time)
emm

# Tukey-adjusted pairwise comparisons across all group × time combinations
contrast(emm, "pairwise", adjust = "tukey")

# 95% confidence intervals for all pairwise contrasts
confint(contrast(emm, "pairwise", adjust = "tukey"))
```

Addressing the primary study questions:
a) Baseline comparison: At Baseline, mean total cholesterol did not differ significantly between the FK506 group (290 mg/dL) and the CsA group (303 mg/dL), with a difference of −13.5 mg/dL (p = 0.934). This confirms that randomization achieved baseline comparability between the two groups.

b) Six-month comparison: At Month 6, the FK506 group had a mean total cholesterol of 233 mg/dL compared to 291 mg/dL in the CsA group, a difference of −58.6 mg/dL (p < 0.001). This difference is clinically significant, exceeding the pre-specified threshold of 40 mg/dL, and indicates that FK506-treated patients achieved meaningfully lower cholesterol levels by six months.

c) Change from baseline to six months: The FK506 group decreased by 56.9 mg/dL from Baseline to Month 6 (p < 0.001), while the CsA group decreased by only 11.9 mg/dL (p = 0.876). The differential change between groups was 45.1 mg/dL, again surpassing the 40 mg/dL clinical threshold, confirming that conversion from CsA to FK506 resulted in a clinically significant reduction in total cholesterol

---

## Step 11: Visualization of Estimated Marginal Means

```{r emm-plot}
emm_df <- as.data.frame(emm)

# Plot model-adjusted means with 95% confidence intervals
ggplot(emm_df, aes(x = time, y = emmean, color = group, group = group)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  scale_color_manual(values = c("FK506 (Treatment)" = "brown",
                                "CsA (Control)"     = "darkgreen")) +
  labs(
    title = "Model-Adjusted Mean Total Cholesterol by Group and Time",
    x     = "Time Point",
    y     = "LS Mean Total Cholesterol (mg/dL)",
    color = "Treatment Group"
  ) +
  theme_minimal() +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )
  
```

---

## Step 12: Model Diagnostics

```{r diagnostics}
residuals_final <- residuals(final_model)
fitted_final    <- fitted(final_model)

# Residuals vs Fitted Values
# Checks the assumption of constant variance (homoscedasticity)
# Residuals should be randomly scattered around zero with no pattern
plot(fitted_final, residuals_final,
     main = "Residuals vs Fitted Values",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch  = 16, col = "steelblue")
abline(h = 0, col = "red", lwd = 2)
grid(col = "gray", lty = "dotted")

# Q-Q Plot of Residuals
# Checks the normality assumption — points should follow the diagonal reference line
qqnorm(residuals_final,
       main = "Q-Q Plot of Residuals",
       pch  = 16, col = "black", cex = 0.7)
qqline(residuals_final, col = "red", lwd = 2)
grid(col = "gray", lty = "dotted")

# Shapiro-Wilk Test
# Formal test of normality — p > 0.05 indicates no significant departure from normality
shapiro.test(residuals_final)
```
